Active-contours driven registration of diffusion MR images of the brain

### 2015-03-14 Publications/2015-NeuroImage/03-introduction.tex
updating intro

% -*- root: 00-main.tex -*-
\section{Introduction}%
\label{sec:introduction}
Image registration is the process to find a spatial mapping that aligns the information available
  in two different coordinate systems.
In their comprehensive survey, \cite{sotiras_deformable_2013} illustrated the wide variety
  of existing methodologies, and classified them by their three principal building blocks:
  the matching criteria to evaluate the proximity of the solution,
  the deformation model that may implement theoretical properties of the mapping
  (i.e. linearity, elasticity, viscosity, etc.), and the optimization method.
The matching criteria can be defined as a function on the intensities of voxels of the
  two images, the alignment of spatial features derived from the corresponding objects in
  the spaces to be registered, or use both intensity and spatial informations.
In \citep{sotiras_deformable_2013} these methods are referred to as \emph{iconic}, \emph{geometric},
  and \emph{hybrid}, respectively.
When segmentation is introduced to define the matching criteria, we face a \emph{hybrid}
  registration method that simultaneously defines a partition of the reference space to identify
  relevant objects in it.
An early integration of segmentation and registration by \cite{bertalmio_morphing_2000} proposed
  a sequential deformation of active contours for object tracking in sequential 2D images.
Shortly, \cite{yezzi_variational_2003} presented the first method including a full solution to
  the registration problem with an affine transformation supporting the coordinates mapping.
\cite{vemuri_joint_2003} proposed an atlas-based registration framework using level sets and only
  one \gls*{pde} for first time.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  extended the ``two \glspl*{pde}'' method of \cite{yezzi_variational_2003}
  to nonlinear registration implementing a free deformation field.
\cite{droske_mumfordshah_2009} reviewed the latter set of techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration and segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
\cite{greve_accurate_2009} presented a widely used registration method called \emph{bbregister},
  included in \emph{FreeSurfer} \citep{fischl_freesurfer_2012}.
Their framework performs robust registration of brain surfaces into the intensity information
  of a target \gls*{mri}, using an affine transformation and active contours with edges.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method using level sets and a nonlinear elasticity smoother on the
  displacement vector field, which preserves topology even with very large deformations.
Finally, \cite{gorthi_active_2011} extended the existing methodologies using a multiphase
  level-set function for the registration of several active contours, in the application
  of atlas-based segmentation.
Alternatively to the historical use of active contours, some Bayesian approaches
  have been proposed as well \citep{wyatt_map_2003,pohl_bayesian_2006,gass_simultaneous_2014}.

We propose \emph{regseg}, a registration method based on active contours without edges
  \citep{chan_active_2001} to drive registration and segment the target image.
The hypothesis underpinning our research is that image registration can be reliably performed
  by searching for homogeneous regions in the target image that correspond to precise contours
  from an atlas, or extracted from other image (i.e. a different time step).
We demonstrate the approach in correcting \gls*{dmri} for nonlinear distortions, an application
  that is described in \autoref{sec:distortions}.
In \autoref{sec:methods_map}, we start from a probabilistic framework to finally show the duality
  between the methods described previously.
Following sections introduce
  the specific features of \emph{regseg} (\autoref{sec:numerical_implementation}),
  state the evaluation experiments (\autoref{sec:experiments_evaluation}),
  and describe the synthetic and real data involved in this study (\autoref{sec:datasets}).
In \autoref{sec:results} we describe our findings to support the usefulness of \emph{regseg}
  in the described application, compared to a well-established tool.
The interpretation of the results, and prospects, are addressed in \autoref{sec:discussion}.

>>>>

% -*- root: 00-main.tex -*-
\section{Introduction}%
\label{sec:introduction}
Image registration is the process to find a spatial mapping that aligns the information available
  in two different coordinate systems.
In their comprehensive survey, \cite{sotiras_deformable_2013} illustrated the wide variety
  of existing methodologies, and classified them by their three principal building blocks:
  the matching criteria to evaluate the proximity of the solution,
  the deformation model that may implement theoretical properties of the mapping
  (i.e. linearity, elasticity, viscosity, etc.), and the optimization method.
The matching criteria can be defined as a function on the intensities of voxels of the
  two images, the alignment of spatial features derived from the corresponding objects in
  the spaces to be registered, or use both intensity and spatial informations.
In \citep{sotiras_deformable_2013} these methods are referred to as \emph{iconic}, \emph{geometric},
  and \emph{hybrid}, respectively.
When segmentation is introduced to define the matching criteria, we face a \emph{hybrid}
  registration method that simultaneously defines a partition of the reference space to identify
  relevant objects in it.
An early integration of segmentation and registration by \cite{bertalmio_morphing_2000} proposed
  a sequential deformation of active contours for object tracking in series of 2D images.
Shortly, \cite{yezzi_variational_2003} presented the first method including a full solution to
  the registration problem with an affine transformation supporting the coordinates mapping.
\cite{vemuri_joint_2003} proposed an atlas-based registration framework using level sets and only
  one \gls*{pde} for first time.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  extended the ``two \glspl*{pde}'' method of \cite{yezzi_variational_2003}
  to nonlinear registration implementing a free deformation field.
\cite{droske_mumfordshah_2009} reviewed the latter set of techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration and segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
\cite{greve_accurate_2009} presented a widely used registration method called \emph{bbregister},
  included in \emph{FreeSurfer} \citep{fischl_freesurfer_2012}.
Their framework performs robust registration of brain surfaces into the intensity information
  of a target \gls*{mri}, using an affine transformation and active contours with edges.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method using level sets and a nonlinear elasticity smoother on the
  displacement vector field, which preserves topology even with very large deformations.
Finally, \cite{gorthi_active_2011} extended the existing methodologies using a multiphase
  level-set function for the registration of several active contours, in the application
  of atlas-based segmentation.
Alternatively to the historical use of active contours, some Bayesian approaches
  have been proposed as well \citep{wyatt_map_2003,pohl_bayesian_2006,gass_simultaneous_2014}.

We propose \emph{regseg}, an active-contours driven registration method.
Then, we demonstrate its aptness on the correction of the nonlinear distortions typically
  presented by \gls*{dmri} data of the brain.
This problem is briefly described in \autoref{sec:distortions}.
The general overview of our method, along with the contributions with respect 
  to the current state of art are summarized in \autoref{sec:contributions}.
In \autoref{sec:methods_map}, we start from a probabilistic framework to finally show the duality
  between the methods described previously.
Following sections
  introduce the specific features of \emph{regseg} (\autoref{sec:numerical_implementation}),
  state the evaluation experiments (\autoref{sec:experiments_evaluation}),
  and describe the synthetic and real data involved in this study (\autoref{sec:datasets}).
In \autoref{sec:results} we describe our findings to support the usefulness of \emph{regseg}
  in the described application, compared to a well-established tool.
The interpretation of the results, and prospects, are addressed in \autoref{sec:discussion}.

### 2015-03-14 Publications/2015-NeuroImage/04-methods.tex
updating intro

  \begin{equation}
  \vec{r}' = \vec{r} + \frac{\gamma \, T_{acq}\, s_{PE}}{2\pi}\Delta B_0(\vec{r}) \cdot \hat{\vec{e}}_{PE},
  \label{eq:fieldmap}
  \end{equation}
%
where $\gamma$ is the gyromagnetic ratio, $T_{acq}$ is the readout time,
  $s_{PE}$ is the pixel spacing along \gls*{pe}, and $\hat{\vec{e}}_{PE}$ the unitary
  vector along \gls*{pe}.
Several methodologies have been proposed to estimate $U_s$ and fix the artifact.
The first solution consisted on estimating $\Delta B_0$ with field mapping
  techniques \citep{andersson_modeling_2001}.
A second family of methods require an extra \gls*{epi} acquisition, but switching
  the \gls*{pe} axis \citep{chiou_simple_2000} or reversing the direction of gradient \gls*{pe}
  increments \citep{cordes_geometric_2000,holland_efficient_2010}.
\cite{kybic_unwarping_2000} proposed a correction using nonlinear registration of the \emph{b0}
  image\footnote{The term \emph{b0} refers to one or more images within the
  \gls*{dmri} dataset that are acquired with a null or very low gradient intensity or
   \emph{b}-value used as signal reference.} of the \gls*{dmri} versus a \gls*{t2} image.
Further developments \citep{irfanoglu_susceptibility_2011} extended this method to the
  three-dimensional problem.
Therefore, a typical protocol for the extraction of structural connectivity
  comprehends \gls*{dmri}, \gls*{t1}, and the acquisition determined by the correction
  method of choice \citep{daducci_connectome_2012}.

\subsection{Our method and contributions}
\label{sec:our_method}
% \emph{Regseg} is related to \emph{bbregister}, as both methods exploit the precise
%   surfaces that can be extracted with \emph{FreeSurfer} \citep{fischl_freesurfer_2012},
%   to then map this prior information into a target image.
% However, \emph{bbregister} features an affine transform that is unable to compensate the
%   distortions.
% To avoid this problem, the active contours are dismissed in regions typically affected by
%   warpings.
% Additionally, \emph{bbregister} is robust in finding clear tissue boundaries.
% In the case of \gls*{dmri} data, oftentimes the registration process is restricted to finding
%   the pial surface, as it is the only clear boundary.
% 
% to the methods based on nonlinear registration of the
%   to the \gls*{t2} image, since we also render the correction as a
%   registration problem.
% Conversely, the method does not require any of the previously mentioned extra acquisitions.
% A fundamental antecedent to \emph{regseg} is found in .
% They proposed a registration method to align surfaces extracted from the \gls*{t1}
%   to \gls*{dmri} images using a linear transformation model using active contours
%   with edges.
% Similarly, we  warped data, in which the scale of the
%   imaged structures is in the range of the image resolution or above.
% Our approach differs from \citep{greve_accurate_2009} in two fundamental choices.
% First, we use a nonlinear deformation model that enables the correction of the
%   susceptibility-derived distortions \citep{jezzard_correction_1995}.
% \cite{greve_accurate_2009} circumvented the nonlinearity by identifying typically
%   distorted regions and removing them from the metric computation.
% Second, our matching criteria does not necessarily need for clear intensity
%   steps at the tissue interfaces where surfaces must be fitted, thanks to the use of
%   active contours without edges \citep{chan_active_2001}.
% This choice is similar to \citep{guyader_combined_2011}, but \emph{regseg} differs in two
%   features.
% First, the regularization relies on the inherent smoothness of the deformation field,
%   and on the anisotropic approach proposed by \cite{nagel_investigation_1986}.
% Second, it accepts several contours to be mapped as in \citep{gorthi_active_2011}.
% However, \emph{regseg} makes use of the explicit representation of surfaces, and
%   does not implement multiphase level sets.
% Our registration method is demonstrated here as a solution to correct for the susceptibility
%   distortions that \gls*{dmri} of the head typically present \citep{jezzard_correction_1995}.
% The distortion correction and the precise segmentation of \gls*{dmri} data is crucial in
% 	applications such as connectome extraction \citep{craddock_imaging_2013}, since the
%   definition of the nodes in the network is generally derived from a parcellation performed in
%   an anatomical \gls*{mri} of the same subject \citep{daducci_connectome_2012}.

@@

\paragraph*{Deformation model}\label{sec:deformation_model}
Let us denote $\{\vec{v}_i\}_{i=1 \ldots N_c}$ the vertices of one or several prior
  surface(s).
In our application, these surfaces are triangularized meshes extracted using \emph{FreeSurfer}
  \citep{fischl_freesurfer_2012}.
The transform $\hat{U}$ \eqref{eq:transform} is supported by a dense deformation field
  $\vec{u}$, such that:

  \begin{equation}
  \vec{v}_i' = \hat{U}\{\vec{v}_i\} = \vec{v}_i + u(\vec{v}_i).
  \label{eq:nodes_tfm}
  \end{equation}

Since the nodes of the anatomical surfaces likely lay off-grid, it is required to
  derive $\vec{u}$ from a discrete set of parameters $\{\vec{u}_k\}_{k=1 \ldots K}$.
Densification is achieved through a set of associated basis functions $\psi_k$:

  \begin{equation}
  \vec{u} = \sum_k \psi_k(\vec{r}) \vec{u}_k.
  \label{eq:intp_kernel}
  \end{equation}

@@

  \begin{equation}
  \frac{\partial u(\vec{r},t)}{\partial t} \propto - \frac{\partial E(\vec{u})}{\partial \vec{u}_k},
  \label{eq:general_gradient_descent}
  \end{equation}
%
  with $t$ being an artificial time parameter of the contour
  evolution, and $\vec{u}_k$ the parameters supporting the estimate
  $\hat{U}$ of the transformation at the current time point.
Now, we introduce \eqref{eq:energy} in \eqref{eq:general_gradient_descent}:

@@

\paragraph*{Real datasets} %
\label{sec:human_connectome}
%
For the evaluation of the algorithm on real \gls*{dmri} data of human brains,
  we collected 16 datasets from the ``minimally preprocessed''
	 database of the \gls*{hcp}.
\todo[inline]{This is partially true: I am currently processing the data and
only 5 subjects with regseg and 10 with the competing method are processed.
All the results presented afterwards are based on this processed cases, and will be
updated for the final submission.}
We refer the reader to \citep{essen_human_2012} for exact details about acquisition
  parameters, and \citep{glasser_minimal_2013} for the preprocessing issues.
The datasets comprehend a large set of images, containing \gls*{t1}, \gls*{t2} and
  multi-shell \gls*{dmri} images.
The original acquisitions are released within ``unprocessed'' packages, and
  the ``minimally preprocessed'' are corrected for artifacts, brain-extracted
  and spatially normalized, along with some results of the standard processing
  pipeline of \emph{FreeSurfer}.

>>>>

  \begin{equation}
  \vec{r}' = \vec{r} + \frac{\gamma \, T_{acq}\, s_{PE}}{2\pi}\Delta B_0(\vec{r}) \cdot \hat{\vec{e}}_{PE},
  \label{eq:fieldmap}
  \end{equation}
%
where $\gamma$ is the gyromagnetic ratio, $T_{acq}$ is the readout time,
  $s_{PE}$ is the pixel spacing along \gls*{pe}, and $\hat{\vec{e}}_{PE}$ the unitary
  vector along \gls*{pe}.
Several methodologies have been proposed to estimate $U_s$ and fix the artifact.
The first solution consisted on estimating $\Delta B_0$ with field mapping
  techniques \citep{andersson_modeling_2001}.
A second family of methods require an extra \gls*{epi} acquisition, but switching
  the \gls*{pe} axis \citep{chiou_simple_2000} or reversing the direction of gradient \gls*{pe}
  increments \citep{cordes_geometric_2000,holland_efficient_2010}.
\cite{kybic_unwarping_2000} proposed a correction using nonlinear registration of the \emph{b0}
  image\footnote{The term \emph{b0} refers to one or more images within the
  \gls*{dmri} dataset that are acquired with a null or very low gradient intensity or
   \emph{b}-value.} of the \gls*{dmri} versus a \gls*{t2} image.
Further developments \citep{irfanoglu_susceptibility_2011} extended this method to the
  three-dimensional problem.
Therefore, a typical protocol for the extraction of structural connectivity
  comprehends \gls*{dmri}, \gls*{t1}, and the acquisition determined by the correction
  method of choice \citep{daducci_connectome_2012}.

\subsection{Our method and contributions}
\label{sec:contributions}
%
\emph{Regseg} can be viewed as a derivation of \emph{bbregister} \citep{greve_accurate_2009}.
Both methods exploit the anatomically correct and precise surfaces that are readily
  extracted with \emph{FreeSurfer} \citep{fischl_freesurfer_2012}, mapping them
  to a target \gls*{dmri} image using active contours.
We implemented active contours without edges \citep{chan_active_2001}, that allow
  \emph{regseg} to segment homogeneous regions within the brain, while \emph{bbregister}
  use active contours with edges to look for intensity boundaries in the target image.
Another identifying difference is the deformation model:
  \emph{bbregister} includes an affine transformation and overcomes the distortion
  problem by excluding typically affected regions from computation.
Conversely, \emph{regseg} implements a free-deformation field supported with B-Spline
  functions.
Our method is close to \citep{guyader_combined_2011} and \citep{gorthi_active_2011}
  in the use of active contours without edges for registration, however we avoided
  the use of level sets with the application of shape gradients
  \citep{besson_dream2s_2003,herbulot_segmentation_2006}.
\cite{gorthi_active_2011} used a gaussian regularizer, and \cite{guyader_combined_2011}
  proposed a nonlinear elasticity smoother.
In contrast, \emph{regseg} implements the anisotropic regularization proposed by
  \citep{nagel_investigation_1986}.
We also present an ellaborated evaluation instrumentation to create realistic distortions
  and test algorithms.
Integrating \emph{regseg} and an in-house implementation of the \gls*{t2b} correction method
  in the workflow, we compared the performance of both solutions.

% \emph{Regseg} is related to \emph{bbregister}, as both methods exploit the precise
%   surfaces that can be extracted with \emph{FreeSurfer} \citep{fischl_freesurfer_2012},
%   to then map this prior information into a target image.
% However, \emph{bbregister} features an affine transform that is unable to compensate the
%   distortions.
% To avoid this problem, the active contours are dismissed in regions typically affected by
%   warpings.
% Additionally, \emph{bbregister} is robust in finding clear tissue boundaries.
% In the case of \gls*{dmri} data, oftentimes the registration process is restricted to finding
%   the pial surface, as it is the only clear boundary.
% 
% to the methods based on nonlinear registration of the
%   to the \gls*{t2} image, since we also render the correction as a
%   registration problem.
% Conversely, the method does not require any of the previously mentioned extra acquisitions.
% A fundamental antecedent to \emph{regseg} is found in .
% They proposed a registration method to align surfaces extracted from the \gls*{t1}
%   to \gls*{dmri} images using a linear transformation model using active contours
%   with edges.
% Similarly, we  warped data, in which the scale of the
%   imaged structures is in the range of the image resolution or above.
% Our approach differs from \citep{greve_accurate_2009} in two fundamental choices.
% First, we use a nonlinear deformation model that enables the correction of the
%   susceptibility-derived distortions \citep{jezzard_correction_1995}.
% \cite{greve_accurate_2009} circumvented the nonlinearity by identifying typically
%   distorted regions and removing them from the metric computation.
% Second, our matching criteria does not necessarily need for clear intensity
%   steps at the tissue interfaces where surfaces must be fitted, thanks to the use of
%   active contours without edges \citep{chan_active_2001}.
% This choice is similar to \citep{guyader_combined_2011}, but \emph{regseg} differs in two
%   features.
% First, the regularization relies on the inherent smoothness of the deformation field,
%   and on the anisotropic approach proposed by \cite{nagel_investigation_1986}.
% Second, it accepts several contours to be mapped as in \citep{gorthi_active_2011}.
% However, \emph{regseg} makes use of the explicit representation of surfaces, and
%   does not implement multiphase level sets.
% Our registration method is demonstrated here as a solution to correct for the susceptibility
%   distortions that \gls*{dmri} of the head typically present \citep{jezzard_correction_1995}.
% The distortion correction and the precise segmentation of \gls*{dmri} data is crucial in
% 	applications such as connectome extraction \citep{craddock_imaging_2013}, since the
%   definition of the nodes in the network is generally derived from a parcellation performed in
%   an anatomical \gls*{mri} of the same subject \citep{daducci_connectome_2012}.

@@

\paragraph*{Deformation model}\label{sec:deformation_model}
Let us denote $\{\vec{v}_i\}_{i=1 \ldots N_c}$ the vertices of one or several prior
  surface(s).
In our application, these surfaces are triangularized meshes extracted using \emph{FreeSurfer}
  \citep{fischl_freesurfer_2012}.
The transform $\hat{U}$ \eqref{eq:transform} is supported by a dense deformation field
  $\vec{u} = u(\vec{r})$, such that:

  \begin{equation}
  \vec{v}_i' = \hat{U}\{\vec{v}_i\} = \vec{v}_i + u(\vec{v}_i) = \vec{v}_i + \vec{u}_i.
  \label{eq:nodes_tfm}
  \end{equation}

Since the nodes of the anatomical surfaces likely lay off-grid, it is required to
  derive $u(\vec{r})$ from a discrete set of parameters $\{\vec{u}_k\}_{k=1 \ldots K}$.
Densification is achieved through a set of associated basis functions $\psi_k$:

  \begin{equation}
  u(\vec{r}) = \sum_k \psi_k(\vec{r}) \vec{u}_k.
  \label{eq:intp_kernel}
  \end{equation}

@@

  \begin{equation}
  \frac{\partial u(\vec{r},t)}{\partial t} \propto - \frac{\partial E(R \mid U)}{\partial \vec{u}_k},
  \label{eq:general_gradient_descent}
  \end{equation}
%
  with $t$ being an artificial time parameter of the contour
  evolution, and $\vec{u}_k$ the parameters supporting the estimate
  $\hat{U}$ of the transformation at the current time point.
Now, we introduce \eqref{eq:energy} in \eqref{eq:general_gradient_descent}:

@@

\paragraph*{Real datasets} %
\label{sec:human_connectome}
%
For the evaluation of the algorithm on real \gls*{dmri} data of human brains,
  we collected 16 datasets from the ``minimally preprocessed''
	 database of the \gls*{hcp}.
We refer the reader to \citep{essen_human_2012} for exact details about acquisition
  parameters, and \citep{glasser_minimal_2013} for the preprocessing issues.
The datasets comprehend a large set of images, containing \gls*{t1}, \gls*{t2} and
  multi-shell \gls*{dmri} images.
The original acquisitions are released within ``unprocessed'' packages, and
  the ``minimally preprocessed'' are corrected for artifacts, brain-extracted
  and spatially normalized, along with some results of the standard processing
  pipeline of \emph{FreeSurfer}.
